SMT-Based Generation of Symbolic Automata \thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2019-04-09 2017-Xudong-SMT/ActaInformatica.tex
Redone proof 3

\begin{algorithm}[h]
  \caption{Open Automaton Generation}
  \label{alg1}
\begin{algorithmic}[1]
\Require A pNet $P$ (can be a PLTS, but not a hole)
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` pn ''}
\State Initialize sets $U=\{\mathit{InitState}(P)\}$ and $E=\emptyset$,
for unexplored and explored global states, respectively; $L=\emptyset$ for the resulting OTs;
\While{!$\mathit{isEmpty}(U)$}
	\State Choose $S$ in $U$; remove $S$ from $U$, add $S$ to $E$;
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` Choose $S$ in $U$ ''}
	\State $\mathit{OTs} = \mathit{MakeTransitions}(P, S)$;
	\For {each $\mathit{OT} \in \mathit{OTs}$}
          \State Add $\mathit{OT}$ to $L$;
          \State Let $S'$ be the target state of $\mathit{OT}$;
          add $S'$ to the unexplored states if necessary:
	  \IfThen{($S' \not\in U \cup E$)}{Add $S'$ into $U$;}
	  \EndFor
\EndWhile
\State $\mathit{L1}$=filterUnsatTransitions($\mathit{L,Assignments}$);
\State $\mathit{OA_0}$=makeReachableSubAutomaton($\mathit{InitState}(P),L1)$;
\Repeat $\mathit{OA_n}$=RefineReachableSubAutomaton($\mathit{OA_{n-1}}$)
\Until  $\mathit{OA_n}=\mathit{OA_{n-1}}$;   \hfill $\setminus \setminus$ Fixed point reached
\State \Return $\mathit{OA_n}$;

@@

Before the SAT checking step, it may contain symbolic transitions that
do not have valid ground instantiations. This is not logically
incorrect, but is naturally not an optimal representation. So we prune
these UNSAT transitions: at the end of the open automaton generation,
the predicate of each OT is translated 
into SMTlib assertions (see Section~\ref{section:pruning}), and checked for
satisfiability (function \texttt{filterUnsatTransitions}). 
Then we compute the set of reachable transitions and states to build
the final open automaton (more details in Section~\ref{section:BuildingTheOpenAutomaton}).

@@

@@

Our matching/filtering strategy builds some transitions where
the predicates express incompatible constraints. Even if having an
unsatisfiable (symbolic) transition would not be incorrect,
we choose to minimize the open automaton (\ie its number of
transitions and states), by checking the predicates for satisfiability.
In Figure~\ref{schema:unsat-ot}, we show an 
unsatisfiable open transition 
from the open automaton of our running example.

It shows the case where processes $P$ and $R$ ``try'' to communicate directly, but this is made impossible by the combination of chosen synchronisation vectors and their guards. This
  mismatch is materialised by the predicate
fragment ``$a1 = \delta(x2) \land \forall z. a1 \neq \delta(z)$''. In Section~\ref{section:TranslationToSMTlib} we will show how this is encoded and checked using the SMT engine.

\begin{figure}[h]
        \begin{mathpar}
	  OT_{19}  =
          \inferrule*[myfraction=\reddottedrule]
	{\{\xrightarrow{a1}_P ~~
          \xrightarrow{acc(x2)}_R \} ~~
          [a1 = \delta(x2) \land \forall z. a1 \neq \delta(z)]
        }
	{\ostate{2,0} \xrightarrow {\underline{\delta(x2)}} \ostate{3,0}}
	\end{mathpar}
\caption{One of the unsatisfiable open transitions}  \label{schema:unsat-ot}
\end{figure}

Checking satisfiability requires some symbolic computation
on the action expressions and the predicates, which
may depend on the specific theory of the action algebra datatypes. 
The ``Modulo Theory'' part of SMT solvers is important here, as
the solvers can use specific properties of each action algebra.

Function \texttt{filterUnsatTransitions}, in line 11 of
Algorithm\ref{alg1}, and function \\
\texttt{RefineReachableSubAutomaton}
in line 13, make use of the Z3 engine to implement the
satisfiability checking, for all the OTs computed that far.

@@

The set $L1$ may contain unreachable transitions. To remove those transitions, the last step of the
algorithm builds the final open automaton via a traversal of $L1$,
starting at the initial state $InitState(p)$ as defined in
page~\pageref{def-states}. More precisely, this is a residual
algorithm building a graph from the reachable transitions in $L1$. 
%and states are the set of target states of the transitions. 
Each state in the graph is decorated with a set of its local
variables, and a list of their possible assignments,
gathered during the traversal using the ``Post'' parts of the
OTs.

A further reduction may now be gained by using the information from
these sets of possible assignments: adding this information when
checking satisfiability of transitions leaving a given state may
result in more UNSAT transitions. The set of possible
assignments computed by the algorithm is an over approximation of
possible values of the 
variables of such state, so narrowing it means we get a better
approximation, and potentially more information, which can be used in
the SAT checking of subsequent transitions. In turn, this may reduce the possible
assignments of the corresponding target states. In
Algorithm~\ref{alg1}, we iterate this procedure, each iteration
reducing the set of 
transitions and the sets of assignments; the final open automaton
is the fixed point of this iteration.

In practice, this could be
expensive, and it is not incorrect to leave some UNSAT OTs in the
result. From a  practical point of view, very little will be gained to compute the
fixed point, so we would only perform one step of iteration. Further algorithms 
(e.g. model-checking or equivalence checking) will be able to use the
full information from the predicates and the assignment sets.

@@

But there is more to this. Due to the symbolic nature of the
  transitions, reachability is not as easy to define as in standard
  models, as illustrated by the definition above. By avoiding building
  the OTs coming from some unreachable states, we also avoid adding
  the corresponding assignments to the variables of their target states.
But we \emph{cannot} use the
assignment information in the ``on-the-fly'' SAT checking, because
they cannot be complete before exploring (at least) all the reachable
state space. We first need to compute a first over-approximation of
the reachable state space and assignment sets; this corresponds to
the automaton $OA_0$ in Algorithm~\ref{alg3}, line 15.
So the smart algorithm proceeds in two steps: during the
first step, the SAT checking is done without using any assignment set
information. Then after the completion of this step, we refine the
reachability computation, using the
assignment information as described in the previous section.

\begin{algorithm}[h]
  \caption{``Smart'' Reachable Open Automaton Generation}
  \label{alg3}
\begin{algorithmic}[1]
\Require A pNet $P$ (cannot be a hole)
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` pn ''}
\State Initialize sets $U=\{\mathit{InitState}(P)\}$ and $E=\emptyset$,
for unexplored and explored global states, respectively; $L=\emptyset$ for the resulting OTs;
\While{!$\mathit{isEmpty}(U)$}
	\State Choose $S$ in $U$; remove $S$ from $U$, add $S$ to $E$;
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` Choose $S$ in $U$ ''}
	\State $\mathit{OTs} = \mathit{MakeTransitions}(P, S)$;
	\For {each $\mathit{OT} \in \mathit{OTs}$}
        Check satisfiability of $\mathit{OT}$ using the SMT solver;
        \IfThen {$SAT(\mathit{OT})$}
                {\State \{Add $\mathit{OT}$ to $L$;
                  \State Let $S'$ be the target global state of $\mathit{OT}$
	          \IfThen{($S' \not\in U \cup E$)}{Add $S'$ into $U$;}
                  \}}
	\EndFor
\EndWhile
\State $\mathit{L2}$=filterUnsatTransitions($\mathit{L,Assignments}$);
\State $\mathit{OA_0}$=makeReachableSubAutomaton($\mathit{InitState}(P),L2)$;
\Repeat $\mathit{OA_n}$=RefineReachableSubAutomaton($\mathit{OA_{n-1}}$)
\Until  $\mathit{OA_n}=\mathit{OA_{n-1}}$;
\State \Return $\mathit{OA_n}$;
%\EndIf
\end{algorithmic}  
\end{algorithm}

Algorithm~\ref{alg3} implements this new stategy: in line 5, it runs
the satisfiability check on each OT generated from this global
state. This check is done without using the assignment information, as
discussed above. Only the OTs found SAT from this check will be added
to $L$, and the corresponding target global state added to $U$ if
necessary. This ensures that all explored global states are
effectively reachable by a chain of SAT OTs.

At the end of the residual loop, we do another traversal of the
automaton, this time using the assignment information collected that
far, and compute the final reachable sub-automaton. 
Naturally, this step is only useful if there are some assignments.

The easy case is when the system does not contain assignments to state
variables. This does not mean ``data-independency'': we can still have
guards in the PLTS transitions and in the synchronisation vectors,
making the behaviour depending on the values of input variables. But
the open automaton needs not to collect the set of possible values dealing
with the values of local variables, so we can directly compute the
reachable part of the automaton using the on-the-fly algorithm (line 13).

@@

% \ERIC{Sketch of the proof:}

%    \medskip
%  \noteInEM{I keep below the original proof draft from Xudong, though
%    I'm not sure if parts of it can be reused}
\begin{proof}[Theorem \ref{Th:CorrectnessWithoutAssignments}]

@@

  Let us first define some notations.
 \begin{itemize}
    \item $L0$: the original set of transitions by Algorithm~\ref{alg1}
      (i.e. before line 11), 
    \item $L1_{sat}$:  the original set of transitions by Algorithm~\ref{alg1}, after
      satisfiability check (i.e. after line 11),
    \item $L1_{reach}$: the final set of OTs of Algorithm~\ref{alg1}, as computed
      by line 12 (as $OA_0$). Without assignments the following
      refinement step will do nothing, the fixpoint is already reached.
    \item $L2_{sat}$: the original set of transitions by Algorithm~\ref{alg3}
      (i.e. before line 13), 
    \item $L2_{reach}$: the final set of OTs of Algorithm~\ref{alg3}, as computed
          by line 13. Here also $OA_n = OA_0$.
   \item $States: \overline{OT}\rightarrow\overline{State}$:  the function that
computes the set of target states of a set of OTs.  
  \item $Rename:  L1_{reach} \rightarrow L2_{reach}$:  a function that keeps state names of transitions but renames fresh variables in the transition labels.
\end{itemize}

%% When the pNets contains no assignment, there is no need for the last
%% filtering step, so $L1_{sat}=L1_{reach}$ and $L2_{sat}=L2_{reach}$.

\smallskip
We now prove that $L1_{reach}=L2_{reach}$, which will imply
the equality of the two open automata.

\smallskip
Let $S=\ostate{\overline{s}}=\ostate{\{s_{i}\}^{i\in Leaves(pn)}}$
range over the global states $States(L0)$, 
The only difference between the main loops of the algorithms is that 
Algorithm~\ref{alg3} only retains the satisfiable OTs (lines 6-7), so 
Algorithm~\ref{alg3} computes fewer transitions than
Algorithm~\ref{alg1}. Formally, we have that $L2_{reach}\subseteq  L2_{sat} \subseteq L0$, and as a consequence $States(L2_{reach})\subseteq  States(L0)$, i.e.,\\
\centerline{$\forall \ostate{\overline{u}}\in States(L2_{reach}), \exists\ostate{\overline{t}}\in States(L0), \ostate{\overline{t}}=\ostate{\overline{u}}$}

@@

  Suppose our property is true for distance $n$, and
  suppose $|S|_{L2_{reach}}=n+1$, then there exists a chain
  $S_0\xrightarrow{l_1}S_1\xrightarrow{l_2}...S_{n}\xrightarrow{l_{n+1}}S$
  with all $l_k \in L2_{reach}$. In particular, since
  $L2_{reach}\subseteq L2_{sat}$, we have that $l_{n+1} \in L2_{sat}$ and
  its predicate is satisfiable.

  By induction hypothesis, we get $S_{n} \in States(L1_{reach})$, so during the
  call of $makeReachableSubAutomaton$ in Algorithm~\ref{alg1}, line 12, the OT
  $S_n \xrightarrow{l_{n}}S$ will be checked and found SAT, so
  $S \in States(L1_{reach})$.

  Conversely, if $|S|_{L1_{reach}}=n+1$, we get a similar chain in
  $L1_{reach}$, with the transition $S_n \xrightarrow{l_{n}}S$.  Then $ S_n \in
  States(L1_{reach})$ and $l_n$ is satisfiable. By induction, we have
  $S_n \in States(L2_{reach})=States(L2_{sat})$, so $l_n$ will have
  been checked on the fly, in line 6 of Algorithm~\ref{alg3}. Then
  the transition $S_n \xrightarrow{l_{n}}S$ will be added to $L2_{sat}$, and thus $S \in States(L2_{reach})$.

@@

\paragraph{General case with assignments.}
In the presence of assignments, we need to collect all of them
  before using the assignment information to refine the SAT
  checking. Still, the smart strategy will explore a reduced state
  space, not necessarily the minimal one. In the
  following theorem, both algorithms are considered in their complete
  version, meaning the reachability (with assignments) step is
  iterated up to fixed points.

@@

%\TODO{Formalize the proof}
\begin{proof}
  Define the OT sets $L0, L1_{sat}, L2_{sat}, L2_{reach}$ and the function $States$ as in the proof of Theorem~\ref{Th:CorrectnessWithoutAssignments}, and

  \begin{itemize}
    \item $L1_{reach*}$: the final set of OTs of Algorithm~\ref{alg1}, as computed
        by an iterative application of $makeReachableSubAutomaton$ (line
        14) up to a  fixed point.
      \item $L2_{reach*}$: the final set of OTs of Algorithm~\ref{alg3}, as computed
        by an iterative application of $makeReachableSubAutomaton$ (line
        16) up to a fixed point.
        \item $Assignments : \overline{OT}\ \times State\rightarrow \overline{Assign}$: the function of
        computing the set of assignments of a global state from a given set of OTs.
  \end{itemize}

@@

\smallskip
The structure implicitly computed by both algorithms is a triple $ Aut= \langle \overline{OT}, \overline{State},\\ \overline{Assign}\rangle$.
The assignments set of $OA_1$ is $\{Assignments(L1, S_i)\}^{S_i\in States(L1)}$ and that of $OA_2$ is $\{Assignments(L2_{sat}, S_j)\}^{S_j\in States(L2_{sat})}$. The result triples are:  
\noindent
\begin{align*}
Aut_1 = \langle L1_{sat}, States(L1_{sat}), \{Assignments(L1, S_i)\}^{S_i\in States(L1)}\rangle\ , \\
Aut_2 = \langle L2_{sat}, States(L2_{sat}), \{Assignments(L2_{sat}, S_j)\}^{S_j\in States(L2_{sat})}\rangle \  .
\end{align*}

Based on this triple structure, we define the function $Refine : Aut \rightarrow Aut$ modelling $makeReachableSubAutomaton(S0,OTs)$. It uses the assignments of the previous step to refine the computed reachable space:
\begin{align*}
Refine(Aut) &=\langle L', States(L'), \{Assignments(L', S_i)\}^{S_i\in States(L')}\rangle \\
&=Aut'
\end{align*}
where $L' = L - L_{unsat}$, and $L_{unsat}$ is the set of unsatisfiable transitions checked in this iteration. We have $L'\subseteq L$, and potentially, their target states and corresponding assignments may be reduced: 
\[States_{L'}=\{S_0\}\cup \bigcup States(L'),\]
\[Assignments_{L',States_{L'}}=\bigcup\{Assignments(L', State_{i})\}^{State_{i}\in States_{L'}} \]
\[s.t. |States_{L'}|\le|States_{L}|\wedge|Assignments_{L',States_{L'}}|\le|Assignments_{L,States_{L}}|.\]

In one single iteration, the satisfiability of a transition $\ostate{\tilde{t}}\xrightarrow{l}\ostate{\tilde{t'}}\in L$ is checked through whether 
\[\bigvee_{Assign_i\in Assignments_{L, States_{L}}}(Pred_{l}\wedge Assign_{i})\text{ is true}.\]
More transitions could become unsatisfiable after the reduction of the assignments set, so the function needs to be iterated several times. In each iteration, the number of triples decreases, thus there exists a sequence \[Aut^n = Refine(Aut^{n-1})\quad s.t. |Aut^n| < |Aut^{n-1}|.\]
The iteration terminates if $Aut''$ contains no more unsatisfiable transition, then there is a fixed point $Aut'' = Refine(Aut'')$ and $|Aut''| = |Refine(Aut'')|$.

Let $Aut''_1$ and $Aut''_2$ be two fixed points with $Refine(Aut''_1) = Aut''_1$ and $Refine(Aut''_2) = Aut''_2$. Focusing on the transition sets in $Aut''_1$ and $Aut''_2$, $L1''_{sat}=L1_{reach}$ and $L2''_{sat}=L2_{reach}$, we need to check their equivalence.

First, there is an inclusion relation between the initial assignments sets, 
\begin{align*}
&L2_{sat}\subseteq L1\Rightarrow States_{L2_{sat}}\subseteq State_{L1}\\
\Rightarrow &Assignments_{L2_{sat}, States_{L2_{sat}}}\subseteq Assignments_{L1, States_{L1}}
\end{align*}

>>>>

\begin{algorithm}[h]
  \caption{Open Automaton Generation}
  \label{alg1}
\begin{algorithmic}[1]
\Require A pNet $P$ (can be a PLTS, but not a hole)
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` pn ''}
\State Initialize sets $U=\{\mathit{InitState}(P)\}$ and $E=\emptyset$,
for unexplored and explored global states, respectively; $L=\emptyset$ for the resulting OTs;
\While{!$\mathit{isEmpty}(U)$}
	\State Choose $S$ in $U$; remove $S$ from $U$, add $S$ to $E$;
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` Choose $S$ in $U$ ''}
	\State $\mathit{OTs} = \mathit{MakeTransitions}(P, S)$;
	\For {each $\mathit{OT} \in \mathit{OTs}$}
          \State Add $\mathit{OT}$ to $L$;
          \State Let $S'$ be the target state of $\mathit{OT}$;
          add $S'$ to the unexplored states if necessary:
	  \IfThen{($S' \not\in U \cup E$)}{Add $S'$ into $U$;}
	  \EndFor
\EndWhile
\State $\mathit{L1}$=filterUnsatTransitions($\mathit{L,Assignments}$);
\State $\mathit{OA_1^0}$=makeReachableSubAutomaton($\mathit{InitState}(P),L1)$;
\Repeat $\mathit{OA_1^n}$=RefineReachableSubAutomaton($\mathit{OA_1^{n-1}}$)
\Until  $\mathit{OA_1^n}=\mathit{OA_1^{n-1}}$;   \hfill $\setminus \setminus$ Fixed point reached
\State \Return $\mathit{OA_1^n}$;

@@

@@

Before the SAT checking step, the set $L$ may contain symbolic transitions that
do not have valid ground instantiations. This is not logically
incorrect, but is naturally not an optimal representation. So (line 11) we prune
these UNSAT transitions: at the end of the open automaton generation,
the predicate of each OT is translated 
into SMTlib assertions (see Section~\ref{section:pruning}), and checked for
satisfiability (function \texttt{filterUnsatTransitions}). 
Then we compute (lines 12-14) the set of reachable transitions and states to build
the final open automaton (more details in Section~\ref{section:BuildingTheOpenAutomaton}).

@@

Our matching/filtering strategy builds some transitions where
the predicates express incompatible constraints. Even if having an
unsatisfiable (symbolic) transition would not be incorrect,
we choose to minimize the open automaton (\ie its number of
transitions and states), by checking the predicates for satisfiability.

\begin{example}[Unsat OT]
  \label{schema:unsat-ot}
\begin{mathpar}
	  OT_{19}  =
          \inferrule*[myfraction=\reddottedrule]
	{\{\xrightarrow{a1}_P ~~
          \xrightarrow{acc(x2)}_R \} ~~
          [a1 = \delta(x2) \land \forall z. a1 \neq \delta(z)]
        }
	{\ostate{2,0} \xrightarrow {\underline{\delta(x2)}} \ostate{3,0}}
	\end{mathpar}

Here we display an  
unsatisfiable open transition 
from the open automaton of our running example.
It shows the case where processes $P$ and $R$ ``try'' to communicate directly, but this is made impossible by the combination of chosen synchronisation vectors and their guards. This
  mismatch is materialised by the predicate
fragment ``$a1 = \delta(x2) \land \forall z. a1 \neq \delta(z)$''. In Section~\ref{section:TranslationToSMTlib} we will show how this is encoded and checked using the SMT engine.
\end{example}

Checking satisfiability requires some symbolic computation
on the action expressions and the predicates, which
may depend on the specific theory of the action algebra datatypes. 
The ``Modulo Theory'' part of SMT solvers is important here, as
the solvers can use specific properties of each data type in the action algebra.

Function \texttt{filterUnsatTransitions}, in line 11 of
Algorithm\ref{alg1}, and function \\
\texttt{RefineReachableSubAutomaton}
in line 13, make use of the Z3 engine to implement the
satisfiability checking, for all the OTs computed that far.
For a transition
\[		\inferrule*[myfraction=\reddottedrule]
		{%\{s_i~{\xrightarrow{a_i}}_i ~s_i^{\prime}\}^{i\in I},
			\{\xrightarrow{\beta_j}_j\}^{j\in J}, \Pred\,, \Post}
		{s \xrightarrow {\alpha}s'}
                \]
                we need to build a predicate encoding both the OT predicate, and
the information available on the possible assignments of the initial
state $s$. This is expressed as
\[Pred_{l}\wedge \bigwedge_{v\in vars(s)}\{ \bigvee_{Assign\in
  assigns(v,s)} Assign\}
\]
where each $Assign$ represents one of the possible (symbolic) value of
variable the state variable $v$.

@@

The set $L1$ may contain unreachable transitions. To remove those
transitions, the last step of the  
algorithm (lines 12-14) builds the final open automaton via a traversal of $L1$,
starting at the initial state $InitState(p)$ as defined in
page~\pageref{def-states}. More precisely, this is a residual
algorithm building a graph from the reachable transitions in $L1$. 
%and states are the set of target states of the transitions. 
Each state in the graph is decorated with a set of its local
variables, and a list of their possible assignments,
gathered during the traversal using the ``Post'' parts of the
OTs.

A further reduction may now be gained by using the information from
these sets of possible assignments: adding this information when
checking satisfiability of transitions leaving a given state may
result in more UNSAT transitions. The set of possible
assignments computed by the algorithm is an over approximation of
possible values of the 
variables of such state, so narrowing it means we get a better
approximation, and potentially more information, which can be used in
the SAT checking of subsequent transitions. In turn, this may reduce the possible
assignments of the corresponding target states. In
Algorithm~\ref{alg1}, we iterate this procedure (lines 13-14), each iteration
reducing the set of 
transitions and the sets of assignments; the final open automaton
is the fixed point of this iteration.

In practice, this could be
expensive, and it is not incorrect to leave some UNSAT OTs in the
result. From a  practical point of view, very little will be gained to compute the
fixed point, so we would only perform one step of iteration. Further algorithms 
(e.g. model-checking or equivalence checking) will be able to use the
full information from the predicates and the assignment sets, even if
some spurious OTs are left here.

@@

But there is more to this. Due to the symbolic nature of the
  transitions, reachability is not as easy to define as in standard
  models, as illustrated by the discussion above. By avoiding building
  the OTs coming from some unreachable states, we also avoid adding
  the corresponding assignments to the variables of their target states.
But we \emph{cannot} use the
assignment information in the ``on-the-fly'' SAT checking, because
they cannot be complete before exploring (at least) all the reachable
state space. We first need to compute a first over-approximation of
the reachable state space and assignment sets; this corresponds to
the automaton $OA_2^0$ in Algorithm~\ref{alg3}, line 13.
So the smart algorithm proceeds in two steps: during the
first step, the SAT checking is done without using any assignment set
information (line 5).  Only the OTs found SAT from this check will be added
to $L$, and the corresponding target global state added to $U$ if
necessary. This ensures that all explored global states are
effectively reachable by a chain of SAT OTs.

\begin{algorithm}[h]
  \caption{``Smart'' Reachable Open Automaton Generation}
  \label{alg3}
\begin{algorithmic}[1]
\Require A pNet $P$ (cannot be a hole)
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` pn ''}
\State Initialize sets $U=\{\mathit{InitState}(P)\}$ and $E=\emptyset$,
for unexplored and explored global states, respectively; $L=\emptyset$ for the resulting OTs;
\While{!$\mathit{isEmpty}(U)$}
	\State Choose $S$ in $U$; remove $S$ from $U$, add $S$ to $E$;
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` Choose $S$ in $U$ ''}
	\State $\mathit{OTs} = \mathit{MakeTransitions}(P, S)$;
	\For {each $\mathit{OT} \in \mathit{OTs}$}
        Check satisfiability of $\mathit{OT}$ using the SMT solver;
        \IfThen {$SAT(\mathit{OT})$}
                {\State \{Add $\mathit{OT}$ to $L$;
                  \State Let $S'$ be the target global state of $\mathit{OT}$
	          \IfThen{($S' \not\in U \cup E$)}{Add $S'$ into $U$;}
                  \}}
	\EndFor
\EndWhile
\State $\mathit{L2}$=filterUnsatTransitions($\mathit{L,Assignments}$);
\State $\mathit{OA_2^0}$=makeReachableSubAutomaton($\mathit{InitState}(P),L2)$;
\Repeat $\mathit{OA_2^n}$=RefineReachableSubAutomaton($\mathit{OA_2^{n-1}}$)
\Until  $\mathit{OA_2^n}=\mathit{OA_2^{n-1}}$;
\State \Return $\mathit{OA_2^n}$;
%\EndIf
\end{algorithmic}  
\end{algorithm}

At the end of the residual loop, we do another traversal of the
automaton, this time using the assignment information collected that
far (line 12), and compute the final reachable sub-automaton
iteratively (lines 13-15), in the same way than in algorithm \ref{alg1}. 
Naturally, this step is only useful if there are some assignments.

\medskip
Now we will prove that the smart algorithm computes the same 
open automaton.

The easy case is when the system does not contain assignments to state
variables. This does not mean ``data-independency'': we can still have
guards in the PLTS transitions and in the synchronisation vectors,
making the behaviour depending on the values of input variables. But
the open automaton needs not to collect the set of possible values dealing
with the values of local variables, so the final automaton is $AO_2^0$ as
computed in line 13.

@@

\begin{proof}[Theorem \ref{Th:CorrectnessWithoutAssignments}]

@@

  Let us first define some notations.
 \begin{itemize}
    \item $L0$: the original set of transitions by Algorithm~\ref{alg1}
      (i.e. before line 11), 
    \item $L1_{sat}$:  the original set of transitions by Algorithm~\ref{alg1}, after
      satisfiability check (i.e. after line 11),
    \item $L1_{reach}$: the final set of OTs of Algorithm~\ref{alg1}, as computed
      by line 12 (as $OA_1^0$). Without assignments the following
      refinement step will do nothing, the fixpoint is already reached.
    \item $L2_{sat}$: the original set of transitions by Algorithm~\ref{alg3}
      (i.e. before line 13), 
    \item $L2_{reach}$: the final set of OTs of Algorithm~\ref{alg3}, as computed
          by line 13. Here also $OA_2^n = OA_2^0$.
   \item $States: \overline{OT}\rightarrow\overline{State}$:  the function that
computes the set of target states of a set of OTs.  
  \item $Rename:  L1_{reach} \rightarrow L2_{reach}$:  a function that
    keeps state names of transitions but renames fresh variables in
    the transition labels. In the sequel, transitions $l_i$ will range
    over $L0$ set (and $L1_*$), and $l'_i$ will range over $L2_*$
    sets. Of course renaming preserves satisfiability.
\end{itemize}

\smallskip
We now prove that $L1_{reach}=L2_{reach}$ and
$States(L1_{reach})=States(L2_{reach})$.

By definition of the algorithms, we have:

\centerline{$L1_{reach} \subseteq L1_{sat} \subseteq L0$}
\centerline{$L2_{reach}\subseteq  L2_{sat}$}

\smallskip
Let $S=\ostate{\overline{s}}=\ostate{\{s_{i}\}^{i\in Leaves(pn)}}$
range over the global states $States(L0)$, 
The only difference between the main loops of the algorithms is that 
Algorithm~\ref{alg3} only retains the satisfiable OTs (lines 6-7), so 
Algorithm~\ref{alg3} computes fewer transitions than
Algorithm~\ref{alg1}. Formally, we have that
$L2_{sat} \subseteq Rename(L0)$ (up to renaming of fresh variables), and as a consequence $States(L2_{reach})\subseteq  States(L0)$, i.e.,\\
\centerline{$\forall \ostate{\overline{u}}\in States(L2_{reach}), \exists\ostate{\overline{t}}\in States(L0), \ostate{\overline{t}}=\ostate{\overline{u}}$}

@@

  Suppose our property is true for distance $n$, and
  suppose $|S|_{L2_{reach}}=n+1$, then there exists a chain
  $S_0\xrightarrow{l'_1}S_1\xrightarrow{l'_2}...S_{n}\xrightarrow{l'_{n+1}}S$
  with all $l'_k \in L2_{reach}$. In particular, since
  $L2_{reach}\subseteq L2_{sat}$, we have that $l'_{n+1} \in L2_{sat}$ and
  its predicate is satisfiable.

  By induction hypothesis, we get $S_{n} \in States(L1_{reach})$, so during the
  call of $makeReachableSubAutomaton$ in Algorithm~\ref{alg1}, line
  12, there exists an OT
  $S_n \xrightarrow{l_{n}}S$ in $L1_{sat}$, with $l'_n=Rename(l_n)$;
  this OT will be checked and found SAT, so
  $S \in States(L1_{reach})$.

  Conversely, if $|S|_{L1_{reach}}=n+1$, we get a similar chain in
  $L1_{reach}$, with the transition $S_n \xrightarrow{l_{n}}S$.  Then $ S_n \in
  States(L1_{reach})$ and $l_n$ is satisfiable. By induction, we have
  $S_n \in States(L2_{reach})=States(L2_{sat})$, so there exists
  $l'_n=Rename(L_n)$ that will have
  been checked on the fly, in line 6 of Algorithm~\ref{alg3}. Then
  the transition $S_n \xrightarrow{l'_{n}}S$ will be added to
  $L2_{sat}$, and thus $S \in States(L2_{reach})$.

@@

\paragraph{General case with assignments.}
In the presence of assignments, we need to collect all of them
  before using the assignment information to refine the SAT
  checking. Still, the smart strategy will explore a reduced state
  space, though not necessarily the minimal one. In the
  following theorem, both algorithms are considered in their complete
  version, meaning the reachability (with assignments) step is
  iterated up to fixed points.

@@

%\TODO{Formalize the proof}
\begin{proof}
  Define the OT sets $L0, L1_{sat}, L2_{sat}, L2_{reach}$ and the
  function $States$ as before, and:

  \begin{itemize}
    \item $L_1^*$: the final set of OTs of Algorithm~\ref{alg1}, as computed
        by an iterative application of $makeReachableSubAutomaton$ (line
        14) up to a  fixed point.
    \item $L_2^*$: the final set of OTs of Algorithm~\ref{alg3}, as computed
        by an iterative application of $makeReachableSubAutomaton$ (line
        15) up to a fixed point.
    \item $Assigns : \overline{OT}\ \rightarrow \overline{Assign}$: the function
          of computing the set of assignments of (the variables of the
          states of) a given set of OTs.
    \item $Assigns : \overline{OT}, Vars\ \rightarrow \overline{Assign}$: similarily,
      but for assignments concerning one specific state variable.
  \end{itemize}

@@

\smallskip
The structure implicitly computed by both algorithms is a triple
$Aut= \langle \overline{OT},\overline{State},\\ \overline{Assign}\rangle$.

We already know that $L2_{sat}\subseteq L1_{sat}$ (up to renaming),
because every OT that will be checked positively (using assignments) in line 12 of
algorithm \ref{alg3} will also be tested by algorithm \ref{alg1} (as
$L2_{sat}\subseteq L0$), and with a larger set of assignments.

Consequently the initial triples of the refinement loop are:

\begin{align*}
Aut_1^0 = \langle L1_{sat}, States(L1_{sat}), Assigns(L1_{sat})\rangle \\
Aut_2^0 = \langle L2_{sat}, States(L2_{sat}), Assigns(L2_{sat})\rangle \\
\text{with } Aut_2^0 \subseteq Aut_1^0, \text{ component-wise and up to renaming.}
\end{align*}

%% The assignments set of $OA_1^0$ is $\{Assigns(L1, States(L1))\}$ and
%% that of $OA_2^0$ is $\{Assigns(L2_{sat}, States(L2_{sat}))}\}$. The
%%   result triples are:    
%% \noindent
%% \begin{align*}
%% Aut_1 = \langle L1_{sat}, States(L1_{sat}), \{Assignments(L1, S_i)\}^{S_i\in States(L1)}\rangle\ , \\
%% Aut_2 = \langle L2_{sat}, States(L2_{sat}), \{Assignments(L2_{sat}, S_j)\}^{S_j\in States(L2_{sat})}\rangle \  .
%% \end{align*}

Based on this triple structure, we define the function $Refine : Aut
\rightarrow Aut$ modelling the function $makeReachableSubAutomaton(S0,OTs)$. It
uses the assignments of the OTs remaining from the previous iteration
to refine the computed reachable space: 
\begin{align*}
Aut^n = Refine(Aut^{n-1}) &=\langle L^n, States(L^n), Assigns(L^n)\rangle
\end{align*}
where $L^n = Reach(L^{n-1} - L^{n-1}_{unsat})$, and $L^{n-1}_{unsat}$ is the
set of unsatisfiable transitions checked in this iteration. We have
$L^n\subseteq L^{n-1}$, and consequently, their states and assignments
may be reduced:
$States(L^n)\subseteq States(L^{n-1}) \land Assigns(L^n)\subseteq
Assigns(L^{n-1})$.

%% \[States_{L'}=\{S_0\}\cup \bigcup States(L'),\]
%% \[Assignments_{L',States_{L'}}=\bigcup\{Assignments(L', State_{i})\}^{State_{i}\in States_{L'}} \]
%% \[s.t. |States_{L'}|\le|States_{L}|\wedge|Assignments_{L',States_{L'}}|\le|Assignments_{L,States_{L}}|.\]

\smallskip
In each iteration $k$, consider a transition $l \in L^k$, define

\[Pred(l,L^{k-1}) = Pred_{l}\wedge \bigwedge_{v\in vars(pn)}\{ \bigvee_{Assign\in
  Assigns(v,L^{k-1})} Assign\}\]

Checking satisfiability of $l$ means checking satifiability of
$Pred(l,L^{k-1})$.

%% \noteInEM{Xudong: Previous version looks wrong to me, because the
%%   disjunction is between the various possible assignments of each
%%   state var, but then you need the conjunction of the information for
%%   all considered vars:
%% \[\bigvee_{Assign_i\in Assignments_{L, States_{L}}}(Pred_{l}\wedge
%% Assign_{i})\text{ is true}.\]}

Note that this is equivalent to the predicate defined in section
\ref{section:pruning}, because we consider all states variables of the
pNet, rather than those of state $s$, but assignments
of other variables will not change the satisfiability.

There are two reasons why an OT would disappear during one application
of $Aut^k = Refine(Aut^{k-1})$: either its predicate, checked with the set of
$Assign(L^{k-1})$, became Unsat; or this OT is at
the end of a sequence starting from a state in $s_0\ \in State(L^k)$,
the next OT became Unsat, and the following states in the sequence are
not reachable using satisfiable OTs in $Aut^k$. More formaly, there is a chain
\[s_0\xrightarrow{ot_1}s_1\xrightarrow{ot_2}...s_{n}\xrightarrow{ot_{n+1}}s_{n+1}\]
with $Pred(ot_1,L^{k-1})$ Unsat, and all states $s_i, i \in [1..n]$
unreachable in $Aut^k$.
